\documentclass[12pt,a4paper]{article}

% ---- Packages ---------------------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}           % professional tables
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{natbib}             % author-year citations
\usepackage{setspace}
\usepackage{microtype}
\usepackage{parskip}            % space between paragraphs, no indent
\usepackage{listings}           % code blocks
\usepackage{caption}

% ---- Code listing style -----------------------------------------------------
\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!8},
  frame=single,
  framerule=0.4pt,
  rulecolor=\color{gray!40},
  breaklines=true,
  breakatwhitespace=true,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  tabsize=4,
  captionpos=b,
  numbers=none,
}
\lstdefinestyle{python}{language=Python}
\lstdefinestyle{bash}{language=bash}

% ---- Hyperref ---------------------------------------------------------------
\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=blue!60!black,
  urlcolor=blue!60!black,
}

% ---- Title block ------------------------------------------------------------
\title{\textbf{gdmbayes}: A Python Package for Bayesian Generalised
       Dissimilarity Modelling}

\author{Harold Horsley\\[4pt]
  \small\textit{Affiliation}\\
  \small Correspondence: \texttt{[email]}
}

\date{}

% =============================================================================
\begin{document}
\maketitle
\doublespacing

% ---- Abstract ---------------------------------------------------------------
\begin{abstract}
Generalised dissimilarity modelling (GDM) is a widely used method for relating pairwise
ecological dissimilarities between sites to environmental and spatial predictors using
non-linear I-spline basis functions. Despite its prevalence in ecology, GDM has until now
lacked a Python implementation and has been confined to a single R package that provides
only point estimates with no uncertainty quantification. A major difficulty in applying GDM
to contemporary ecological datasets lies in the inability to propagate uncertainty through
predictor effects or to incorporate spatial random effects within the existing tooling.

We present \textbf{gdmbayes}, an open-source Python package that fills this gap by
providing both a frequentist scikit-learn-compatible GDM estimator and a full Bayesian
backend built on PyMC\@. The Bayesian estimator supports flexible variance structures,
Gaussian process spatial random effects, and hierarchical predictor importance weights, all
within a reproducible, configuration-driven workflow. \textbf{gdmbayes} follows
scikit-learn conventions throughout, enabling seamless integration with Python's scientific
computing ecosystem.

We demonstrate the package on a simulated species dissimilarity dataset and show that the
Bayesian extension recovers predictor effects with well-calibrated posterior uncertainty.
\textbf{gdmbayes} is the first Python GDM implementation and the only
implementation---in any language---to offer full Bayesian inference for GDM\@. The package
is freely available under the Apache~2.0 licence at
\url{https://github.com/harryhorsley9/spgdmm}.
\end{abstract}

\clearpage

% ---- 1. Introduction --------------------------------------------------------
\section{Introduction}

Ecologists increasingly rely on large spatially referenced datasets to understand how
species composition changes across landscapes. Generalised dissimilarity modelling
\citep[GDM;][]{ferrier2007} has become one of the most widely applied frameworks for this
purpose. GDM relates pairwise ecological dissimilarities---typically Bray--Curtis or
S{\o}rensen distances computed from species assemblage data---to pairwise differences in
environmental predictors and geographic distance using a generalised additive model
structure built on I-spline basis functions. The non-linear I-spline representation allows
GDM to capture the empirical observation that the rate of compositional turnover saturates
at high environmental distances, a biological property that linear models cannot
accommodate.

GDM has been applied across a wide range of ecological contexts including beta-diversity
gradients \citep{ferrier2007}, conservation planning \citep{mokany2012},
climate-change vulnerability assessment \citep{fitzpatrick2015}, and spatial
prioritisation of biodiversity surveys. The R package \textbf{gdm}
\citep{fitzpatrick2021} has been the sole implementation of the method for nearly two
decades, providing a point-estimate frequentist estimator via non-negative least squares
(NNLS). Despite the method's popularity, two important limitations have persisted. First,
there is no Python implementation, which limits integration with the broader scientific
Python ecosystem and with machine-learning pipelines built around scikit-learn. Second,
neither the R package nor any other tool provides uncertainty estimates for GDM
coefficients or predictions, making it impossible to propagate ecological uncertainty
through downstream analyses such as conservation prioritisation or climate projection.

Bayesian inference provides a natural solution to both problems. A Bayesian GDM yields
full posterior distributions over predictor effects, enabling principled uncertainty
quantification for predictor importance scores and for predicted dissimilarities at new
sites. Spatial random effects---structured residual variation unexplained by the available
environmental predictors---can also be incorporated within the Bayesian framework using
Gaussian process priors. These extensions are relevant whenever spatial autocorrelation is
present in species data, as is typically the case in observational ecology.

Here we present \textbf{gdmbayes}, a Python package that provides (1) a frequentist GDM
estimator that replicates the R \textbf{gdm} algorithm and is fully compatible with the
scikit-learn API \citep{pedregosa2011}, and (2) a Bayesian GDM estimator that uses PyMC
\citep{salvatier2016} for MCMC sampling and ArviZ \citep{kumar2019} for posterior
diagnostics. Both estimators share a common preprocessing layer that handles I-spline mesh
construction, pairwise distance computation, and feature matrix assembly. This paper
describes the package architecture, the statistical model, and demonstrates its use on a
simulated dataset.

% ---- 2. Package Description -------------------------------------------------
\section{Package Description}

\subsection{Overview and Installation}

\textbf{gdmbayes} is a Python package for modelling pairwise ecological dissimilarities as
a function of environmental and spatial predictors. The package is organised around three
main classes---\texttt{GDM}, \texttt{spGDMM}, and \texttt{GDMModel}---sharing a common
preprocessing component (\texttt{GDMPreprocessor}). All classes follow the scikit-learn
\texttt{fit(X,~y)} / \texttt{predict(X)} API.

The package is installed via pip:

\begin{lstlisting}[style=bash]
pip install gdmbayes
\end{lstlisting}

A conda environment file is provided for reproducibility. Dependencies include
PyMC~$\geq$~5.0, ArviZ, nutpie (for the NUTS sampler), NumPy, pandas, SciPy, xarray, and
scikit-learn.

The conceptual workflow is:
\begin{enumerate}
  \item Prepare a site-level \texttt{pd.DataFrame} with columns
        \texttt{[xc, yc, time\_idx, predictor\_1, \ldots]} and a condensed pairwise
        dissimilarity vector $\mathbf{y}$ of length $n(n-1)/2$.
  \item Optionally configure preprocessing via \texttt{PreprocessorConfig} and model
        structure via \texttt{ModelConfig}.
  \item Call \texttt{model.fit(X, y)} to fit the frequentist or Bayesian estimator.
  \item Inspect coefficients, predictor importance, or posterior distributions, or generate
        predictions at new sites via \texttt{model.predict(X)}.
  \item Use built-in plotting functions (\texttt{plot\_isplines}, \texttt{plot\_ppc},
        \texttt{plot\_crps\_comparison}) for diagnostics.
\end{enumerate}

\subsection{Preprocessing}

All preprocessing is handled by \texttt{GDMPreprocessor}, a standalone scikit-learn-compatible
transformer (\texttt{BaseEstimator}, \texttt{TransformerMixin}). The preprocessor computes
the I-spline feature matrix from raw site-level data and stores fitted state (knot meshes,
location arrays) for use at prediction time.

\paragraph{I-spline basis functions.}
For each environmental predictor $x$, the preprocessor constructs $J = \texttt{deg} +
\texttt{knots}$ I-spline basis functions $I_1(x), \ldots, I_J(x)$ over a mesh of evenly
spaced or percentile-based knot positions. I-splines are the integral of M-splines and are
monotone non-decreasing by construction \citep{ramsay1988}, which ensures that predicted
dissimilarity increases monotonically with environmental distance. The pairwise feature for
predictor $k$ between sites $i$ and $j$ is:

\begin{equation}
  f_{k,\ell}(i,j) = \bigl|I_\ell(x_{k,i}) - I_\ell(x_{k,j})\bigr|, \quad
  \ell = 1, \ldots, J.
  \label{eq:ispline-feature}
\end{equation}

Geographic distance between sites is optionally included as an additional predictor using
the same I-spline transformation applied to pairwise Euclidean, geodesic, or grid-based
ocean distances. Knot placement is controlled by \texttt{PreprocessorConfig}:

\begin{lstlisting}[style=python]
from gdmbayes import PreprocessorConfig

preprocessor_config = PreprocessorConfig(
    deg=3,                        # I-spline polynomial degree
    knots=2,                      # number of internal knots
    mesh_choice="percentile",     # "percentile", "even", or "custom"
    distance_measure="euclidean", # "euclidean", "geodesic", "ocean_distance"
)
\end{lstlisting}

\subsection{Frequentist GDM}
\label{sec:frequentist}

\texttt{GDM} implements the standard GDM algorithm from \citet{ferrier2007} as a
scikit-learn estimator. Given the pairwise I-spline feature matrix
$\mathbf{X}_{\mathrm{GDM}} \in \mathbb{R}^{P \times Q}$ ($P$ site pairs, $Q$ I-spline
columns), the frequentist estimator solves:

\begin{equation}
  \min_{\boldsymbol{\beta} \geq 0}
  \bigl\| g(\mathbf{y}) - \mathbf{X}_{\mathrm{GDM}}\boldsymbol{\beta} \bigr\|^2,
  \label{eq:nnls}
\end{equation}

where $g(y) = -\log(1 - y)$ is the complementary log-log (cloglog) link applied to the
observed Bray--Curtis dissimilarities $y \in (0, 1)$. Non-negative least squares
\citep[NNLS;][]{lawson1974} enforces $\boldsymbol{\beta} \geq 0$, consistent with the
monotone I-spline representation. Predicted dissimilarities are recovered via the inverse
link:

\begin{equation}
  \hat{y} = 1 - \exp\!\bigl(-\mathbf{X}_{\mathrm{GDM}}\hat{\boldsymbol{\beta}}\bigr).
  \label{eq:inverse-link}
\end{equation}

Predictor importance is defined as the sum of NNLS coefficients over all I-spline basis
functions for each predictor, analogous to the R \textbf{gdm} package. Deviance explained
is the fraction of null deviance (sum of squares of $g(\mathbf{y})$ about its mean)
accounted for by the model.

\begin{lstlisting}[style=python]
from gdmbayes import GDM

model = GDM(geo=True, splines=3, knots=2)
model.fit(X, y)

print(f"Deviance explained: {model.explained_:.4f}")
print(f"Predictor importance: {model.predictor_importance_}")

predictions = model.predict(X_new)  # ndarray of shape (n_pairs,)
\end{lstlisting}

Because \texttt{GDM} inherits from scikit-learn's \texttt{BaseEstimator} and
\texttt{RegressorMixin}, it participates fully in cross-validation, hyperparameter search,
and pipeline composition.

\subsection{Bayesian GDM (\texttt{spGDMM})}
\label{sec:bayesian}

\texttt{spGDMM} (Spatial Generalised Dissimilarity Mixed Model) extends the frequentist
GDM with full Bayesian inference via PyMC. The response variable is modelled on the log
scale:

\begin{equation}
  \log(y_{ij}) \sim \mathrm{CensoredNormal}\!\left(\mu_{ij},\; \sigma^2_{ij},\;
  \text{upper} = 0\right),
  \label{eq:likelihood}
\end{equation}

where the upper censoring at~0 on the log scale (equivalently $y \leq 1$ on the original
scale) reflects the bounded nature of dissimilarities.

\paragraph{Linear predictor.}
With the \texttt{alpha\_importance} option (default), the linear predictor for site pair
$(i, j)$ is:

\begin{equation}
  \mu_{ij} = \beta_0
  + \sum_{k=1}^{K} \alpha_k \sum_{\ell=1}^{J} \beta_{k,\ell}\, f_{k,\ell}(i,j)
  + \sum_{\ell=1}^{J} \beta^\mathrm{d}_\ell\, I_\ell(d_{ij}),
  \label{eq:linear-predictor}
\end{equation}

where $\beta_0 \sim \mathcal{N}(0, 10)$ is a global intercept;
$\boldsymbol{\beta}_k = (\beta_{k,1}, \ldots, \beta_{k,J})
\sim \mathrm{Dirichlet}(\mathbf{1})$ are normalised I-spline coefficients for predictor
$k$; $\boldsymbol{\alpha} = (\alpha_1, \ldots, \alpha_K) \sim
\mathrm{HalfNormal}(1)$ are per-predictor scale parameters (importance weights); and
$\boldsymbol{\beta}^\mathrm{d} \sim \mathrm{LogNormal}(0, 1)$ are geographic distance
coefficients. The Dirichlet prior on $\boldsymbol{\beta}_k$ enforces
$\sum_\ell \beta_{k,\ell} = 1$, so that $\alpha_k$ captures the total magnitude of
predictor $k$'s contribution and $\boldsymbol{\beta}_k$ controls how that contribution is
distributed across the I-spline basis. Without \texttt{alpha\_importance}, coefficients
follow independent $\mathrm{LogNormal}(0, 1)$ priors directly.

\paragraph{Variance structure.}
Three built-in variance functions are available:
\begin{itemize}
  \item \texttt{"homogeneous"}: $\sigma^2 \sim \mathrm{HalfNormal}(1)$, constant across
        all pairs.
  \item \texttt{"covariate\_dependent"}: $\sigma^2$ is a function of pairwise geographic
        distance.
  \item \texttt{"polynomial"}: $\sigma^2$ is a polynomial function of the linear predictor
        $\mu$.
\end{itemize}
Custom variance functions can be passed as Python callables with signature
\texttt{fn(mu, X\_sigma) -> sigma2}, executed inside the PyMC model context.

\paragraph{Spatial random effects.}
An optional Gaussian process prior models spatial autocorrelation in residual
dissimilarities. When \texttt{spatial\_effect} is set, a latent GP

\begin{equation}
  \boldsymbol{\psi} \sim \mathcal{GP}\!\left(0,\;
  \sigma^2_\psi \cdot \mathrm{Exp}\!\left(\|\cdot\| / \ell\right)\right)
  \label{eq:gp-prior}
\end{equation}

is placed on the $n$ training sites. The pairwise spatial effect between sites $i$ and $j$
is added to $\mu_{ij}$. The length scale $\ell$ is set to half the median inter-site
distance; $\sigma^2_\psi \sim \mathrm{InverseGamma}(1,1)$. Two built-in contrast
functions are available (\texttt{"abs\_diff"} and \texttt{"squared\_diff"}), and custom
callables are supported.

\paragraph{Sampler.}
MCMC is performed with the No-U-Turn Sampler \citep[NUTS;][]{hoffman2014} via PyMC, with
nutpie as the default backend for improved sampling speed. Sampler settings are controlled
via \texttt{SamplerConfig}:

\begin{lstlisting}[style=python]
from gdmbayes import spGDMM, ModelConfig, PreprocessorConfig, SamplerConfig
from gdmbayes import summarise_sampling

model = spGDMM(
    preprocessor=PreprocessorConfig(deg=3, knots=2),
    model_config=ModelConfig(variance="homogeneous", spatial_effect="abs_diff"),
    sampler_config=SamplerConfig(draws=1000, tune=1000, chains=4, random_seed=42),
)
idata = model.fit(X, y)

diagnostics = summarise_sampling(idata)  # ESS and R-hat summary
\end{lstlisting}

After sampling, inference data is stored in an ArviZ \texttt{InferenceData} object
(\texttt{model.idata}), enabling the full ArviZ diagnostic toolkit (R-hat, effective
sample size, posterior predictive checks).

\subsection{GDMModel and R-compatible Output}

\texttt{GDMModel} wraps \texttt{spGDMM} and returns a \texttt{GDMResult} dataclass whose
attributes mirror the R \textbf{gdm} package output (\texttt{deviance},
\texttt{nulldeviance}, \texttt{explained}, \texttt{intercept}, \texttt{predictors},
\texttt{coefficients}, \texttt{knots}, \texttt{splines}). This provides a drop-in output
format for users migrating from R:

\begin{lstlisting}[style=python]
from gdmbayes import GDMModel

model = GDMModel(geo=True, splines=3)
result = model.fit(X, y, dataname="my_survey")

print(f"Deviance explained: {result.explained:.1f}%")
print(f"Predictors: {result.predictors}")
\end{lstlisting}

\subsection{Model Serialisation}

Fitted \texttt{spGDMM} models are serialised as NetCDF files via \texttt{model.save(fname)}
and restored via \texttt{spGDMM.load(fname)}. The serialisation stores the full
\texttt{InferenceData} object (posterior samples, prior/posterior predictive samples,
training data) together with the fitted preprocessor state (spline meshes, knot positions,
site coordinates) as xarray datasets in the \texttt{constant\_data} group. This enables
prediction on new data after deserialisation without refitting:

\begin{lstlisting}[style=python]
model.save("fitted_gdm.nc")
model2 = spGDMM.load("fitted_gdm.nc")
predictions = model2.predict(X_new)
\end{lstlisting}

\subsection{Plotting and Diagnostics}

The \texttt{gdmbayes.plotting} module provides four functions:

\begin{itemize}
  \item \texttt{plot\_isplines(model)}: Plot posterior median I-spline effect curves per
        predictor, with optional credible intervals.
  \item \texttt{plot\_ppc(idata, y\_obs)}: Posterior predictive check comparing observed
        and predicted dissimilarity distributions.
  \item \texttt{plot\_crps\_comparison(y\_test, y\_pred, y\_train)}: Compare model CRPS
        against a null (mean) baseline.
  \item \texttt{summarise\_sampling(idata)}: Tabular ESS and R-hat diagnostics for all
        parameters.
\end{itemize}

% ---- 3. Usage Examples ------------------------------------------------------
\section{Usage Examples}

\subsection{Frequentist GDM}

\begin{lstlisting}[style=python, caption={Fitting the frequentist GDM to simulated data.}]
import numpy as np
import pandas as pd
from scipy.spatial.distance import pdist
from gdmbayes import GDM

np.random.seed(42)
n = 30
X = pd.DataFrame({
    "xc":       np.random.uniform(0, 100, n),
    "yc":       np.random.uniform(0, 100, n),
    "time_idx": np.zeros(n),
    "temp":     np.random.uniform(5, 20, n),
    "precip":   np.random.uniform(400, 1200, n),
})
species = np.random.exponential(1, (n, 50))
y = pdist(species, "braycurtis").clip(1e-8, 1 - 1e-8)

model = GDM(geo=True, splines=3, knots=2)
model.fit(X, y)
print(f"Explained: {model.explained_:.3f}")
print(f"Importance: {model.predictor_importance_}")
\end{lstlisting}

\subsection{Bayesian GDM with Spatial Effect}

\begin{lstlisting}[style=python, caption={Fitting the Bayesian spGDMM with a GP spatial effect.}]
from gdmbayes import spGDMM, ModelConfig, PreprocessorConfig, SamplerConfig
from gdmbayes import plot_isplines, plot_ppc

model = spGDMM(
    preprocessor=PreprocessorConfig(deg=3, knots=2, distance_measure="euclidean"),
    model_config=ModelConfig(
        variance="homogeneous",
        spatial_effect="abs_diff",
        alpha_importance=True,
    ),
    sampler_config=SamplerConfig(draws=1000, tune=1000, chains=4, random_seed=42),
)
idata = model.fit(X, y)

plot_isplines(model)   # posterior I-spline curves
plot_ppc(idata, y)     # posterior predictive check
\end{lstlisting}

\subsection{Custom Variance Function}

The Bayesian model structure is extensible via callables. The following defines a variance
that scales exponentially with the linear predictor:

\begin{lstlisting}[style=python, caption={Defining a custom heteroskedastic variance function.}]
import pymc as pm
from gdmbayes import ModelConfig, spGDMM

def heteroskedastic_variance(mu, X_sigma):
    beta_s = pm.HalfNormal("beta_s", sigma=1)
    return beta_s * pm.math.exp(0.5 * mu)

model = spGDMM(
    model_config=ModelConfig(variance=heteroskedastic_variance)
)
idata = model.fit(X, y)
\end{lstlisting}

% ---- 4. Comparison ----------------------------------------------------------
\section{Comparison with the R \textbf{gdm} Package}

Table~\ref{tab:comparison} compares \textbf{gdmbayes} with the R \textbf{gdm} package
\citep{fitzpatrick2021} across key features.

\begin{table}[h]
  \centering
  \caption{Feature comparison of \textbf{gdmbayes} (Python) and \textbf{gdm} (R).}
  \label{tab:comparison}
  \begin{tabular}{lcc}
    \toprule
    Feature & \textbf{gdmbayes} (Python) & \textbf{gdm} (R) \\
    \midrule
    Frequentist GDM (NNLS)                  & Yes            & Yes \\
    Bayesian inference (MCMC)               & Yes            & No  \\
    Posterior uncertainty for coefficients  & Yes            & No  \\
    Posterior uncertainty for predictions   & Yes            & No  \\
    Spatial random effects (GP)             & Yes            & No  \\
    Custom variance structures              & Yes            & No  \\
    Predictor importance weights            & Yes            & Yes \\
    I-spline basis (monotone)               & Yes            & Yes \\
    Geodesic distance                       & Yes            & Partial \\
    sklearn API (\texttt{fit}/\texttt{predict}) & Yes        & No  \\
    Pipeline composition                    & Yes            & No  \\
    Model serialisation                     & Yes (NetCDF)   & Yes (R object) \\
    Posterior predictive checks             & Yes (ArviZ)    & No  \\
    Language                                & Python $\geq$ 3.10 & R \\
    \bottomrule
  \end{tabular}
\end{table}

The R \textbf{gdm} package remains the reference implementation for frequentist GDM and
provides additional utilities such as partial regression plots and Monte Carlo significance
testing that are not yet included in \textbf{gdmbayes}. \textbf{gdmbayes} prioritises
Bayesian inference, uncertainty quantification, and Python ecosystem integration.

% ---- 5. Availability --------------------------------------------------------
\section{Availability}

\textbf{gdmbayes} is available at \url{https://github.com/harryhorsley9/spgdmm} under the
Apache~2.0 licence. A PyPI release is forthcoming. A Zenodo DOI will be minted at the time
of publication. The package requires Python~$\geq$~3.10 and is tested on Linux.

% ---- Acknowledgements -------------------------------------------------------
\section*{Acknowledgements}

[To be completed.]

% ---- Author Contributions ---------------------------------------------------
\section*{Author Contributions}

H.~Horsley: conceptualisation, software, formal analysis, writing -- original draft,
writing -- review and editing.

% ---- Data Availability ------------------------------------------------------
\section*{Data Availability}

No empirical data were used in this paper. Example scripts in the \texttt{examples/}
directory of the repository are fully self-contained and generate synthetic data.

% ---- References -------------------------------------------------------------
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
